---
title: 'PHIL 444: Game Theory and Social Choice'
author: "Brian Weatherson"
date: "Fall 2026"
format:
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    fig-cap-location: bottom
    geometry: margin=1in
    # EB Garamond Math is a custom build with the maths characters added,
    # used for both text and maths. It has no bold or italic, so those come
    # from the stock EB Garamond faces. Same setup as notes/_quarto.yml.
    mainfont: "EB Garamond Math"
    mainfontoptions:
      - BoldFont=EB Garamond SemiBold
      - ItalicFont=EB Garamond Italic
      - BoldItalicFont=EB Garamond SemiBold Italic
    mathfont: "EB Garamond Math"
    date-format: "[Fall] YYYY"
    include-in-header:
      text: |
        \linespread{1.05}
        \RedeclareSectionCommand[
          afterskip=0.3ex,
          beforeskip=0.8ex
          ]{subsection}
        \RedeclareSectionCommand[
          afterskip=0.1ex,
          beforeskip=0.5ex
          ]{subsubsection}
  html:
    theme: cosmo
    code-tools: false
    axe: true
    toc: true
    shift-heading-level-by: 1
    date-format: "[Fall] YYYY"
  docx: default
---

```{r}
#| label: setup
#| include: false
#| cache: false
#| message: false
#| warning: false

knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(fig.path = 'figs/')
knitr::opts_chunk$set(cache.path = '_cache/')
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
```

```{r}
#| label: dates_bib
#| include: false
#| cache: false

mon <- as.Date("2026-08-31") # Monday of the First Week (per Fall 2026 academic calendar)
tue <- mon + 1
wed <- tue + 1
thu <- wed + 1
fri <- thu + 1

advdate <- function(obj, adv) {
 theday <- obj + 7*(adv-1)
 ttheday <- format(theday, format="%A, %B %d")
 return(ttheday)
}
```

| **Lead Instructor**: Brian Weatherson
| **Email**: weath@umich.edu
| **Web**: canvas.umich.edu
| **Office Hours**: TBD

&nbsp;

# Key Points

- This course examines two philosophical questions about choice in social settings.
- The first half is on game theory: what is rational choice when the unknowns are the actions of other rational agents?
- The second half works through Amartya Sen's *Collective Choice and Social Welfare* (2017 expanded edition): what does it mean for a group to choose, and how can individual judgments be combined into collective ones?
- The mathematics is more demanding than a typical philosophy course but does not require calculus. The level is roughly that of formal work in a philosophy journal.
- Assessment consists of two essays (25% each), eight quizzes with the best six counting for 40% in total, and class participation (10%).

# Course Description

This course examines two philosophical questions about choice in social settings.

The first half asks what rational choice looks like when the unknowns are not natural events (the weather, the bus arrival time) but the actions of other rational agents who are reasoning about you. Standard expected-utility theory does not pin down behavior in such settings. We work through the central solution concepts of game theory (Nash equilibrium, rationalizability, subgame perfection, perfect Bayesian equilibrium) and ask what philosophical claims each one rests on. The thread builds toward signaling games and equilibrium refinements, the beer-quiche example and the Cho-Kreps intuitive criterion, with applications to lemons markets and job-market signaling.

The second half asks what it means for a group to choose. Individually rational preferences do not always combine into a coherent group preference: this is the lesson of Arrow's theorem, of Sen's Paretian liberal, and of related impossibility results. Working from Sen's *Collective Choice and Social Welfare* (2017 expanded edition), we study the formalism behind these results and the philosophical responses Sen develops, including his work on interpersonal comparison, justice, rights, and the limits of welfarism.

The unifying theme is that individual rationality does not extend cleanly to social settings. Game theory and social choice theory are two interestingly different ways of seeing how, and what philosophers have made of it.

# Required Materials

The required text is:

- Amartya Sen, *Collective Choice and Social Welfare: An Expanded Edition* (Harvard University Press, 2017). Available electronically through the University of Michigan library.

The main reading for the first half is my own lecture notes, *Game Theory and Social Choice*, free at <https://bweatherson.github.io/F26-Phil444-site/notes/>, supplemented by three classic papers (Akerlof on lemons, Spence on job-market signalling, Cho and Kreps on the intuitive criterion). Giacomo Bonanno's *Game Theory*, also free, is a good second voice if a topic is not landing.

Three readings in the second half come from outside CCSW: Sen's Nobel lecture "The Possibility of Social Choice" (1999), Sen's "Equality of What?" (1979), and Nussbaum's "Capabilities as Fundamental Entitlements" (2003). Links are given in the schedule below. Most of them will only resolve from a campus connection or through the library proxy.

All readings will be made available through Canvas or the course website.

# Course Requirements

- There will be 2 essays, each worth 25% of the grade, one for each half of the course.
- There will be 8 quizzes over the semester. Your best 6 results count, and together they are worth 40% of the grade.
- 10% of the grade is for participation in class.

# Summary of Grading System

1. Essays - 2 essays, 25% each: 50%
2. Quizzes - best 6 of 8: 40%
3. Class participation: 10%

# Canvas

There is a Canvas site for this course, which can be accessed from <https://canvas.umich.edu>. Course documents will be available from this site. Please make sure that you can access this site. Consult the site regularly for announcements, including changes to the course schedule. And there are many tools on the site to communicate with each other, and with me.

There is also a github site for the course which contains more documents, and it is at <https://bweatherson.github.io/F26-Phil444-site/>.

\pagebreak

# Class Schedule

Readings should be done before the scheduled class. Lectures 1-14 cover game theory; Lectures 15-28 cover social choice theory, working through Sen's *Collective Choice and Social Welfare* (cited below as CCSW). "Notes" below refers to my lecture notes, *Game Theory and Social Choice*, at <https://bweatherson.github.io/F26-Phil444-site/notes/>.

CCSW alternates between chapters and starred chapters. The unstarred chapters make the argument in prose; the starred chapters contain the proofs. Sen says in his 1970 preface that the unstarred chapters can be read on their own, and this course takes him at his word. **Unstarred chapters are required. Starred chapters are proved at the board and listed as recommended.** Three starred chapters are exceptions, because in those cases the starred chapter is the lecture: Ch 2* on collective choice rules, Ch 6* on the liberal paradox, and Ch A1* on Arrow's theorem.

Nothing formal is being skipped by this arrangement. The proofs happen in class; the starred chapters are where to find them written down. Essay topics will often be drawn from the recommended reading, so the reading you do for the essay you are actually writing should go well beyond this list.

**How firm the second half is.** Lectures 15-28 work through a single book, and how fast a seminar gets through a book depends on the seminar. What follows is what I expect to do, not a promise. Where a topic listed for one class needs two, it will get two.

The order will not change. Lectures 16, 17, 18 and 21 build the apparatus that everything later uses, and nothing will be skipped or resequenced to make up time. If time has to come from somewhere, it comes from the end of the list, in this order: Lectures 27 and 28 merge into one class, and then Lecture 24 folds into Lecture 23. Lecture 26 stays where it is regardless.

Changes to a week's reading will be posted on Canvas by the Friday before. If nothing is posted, what is below stands.

Quizzes cover the material through the previous class meeting, whatever that turned out to be, rather than whatever this schedule says it should have been.

## Week 1: Decision and Strategic Settings

### `r advdate(tue, 1)` (Lecture 1)

**Topic**:    Course overview and the two organizing questions. A first pass at expected value, and the puzzle that arises when the "states" are other rational agents. Motivating examples: a coordination problem and the Prisoner's Dilemma.    
**Reading**:    None.

### `r advdate(thu, 1)` (Lecture 2)

**Topic**:    Decision theory under risk and uncertainty: expected utility, Allais and Ellsberg. The formal contrast with strategic settings.    
**Reading**:    Notes, Ch 2 (Probability and Decision Theory), 2.4-2.5 (Expected Value; Orthodox Decision Theory).    
**Recommended**:    Notes, 2.1-2.3, if you want the probability revision.

## Week 2: Strategic Form Games

### `r advdate(tue, 2)` (Lecture 3)

**Topic**:    Strategic form games; dominance and iterated elimination of dominated strategies.    
**Reading**:    Notes, Ch 1 (Basics of Game Theory), all of it. The four examples in 1.3-1.7 are the ones we used in Lecture 1; dominance is 1.8-1.10.    
**Recommended**:    Notes, 3.7 (Iterated Dominance), which we return to in Lecture 4.

### `r advdate(thu, 2)` (Lecture 4)

**Topic**:    Nash equilibrium.    
**Reading**:    Notes, Ch 3 (Mixed Strategies and Nash Equilibrium), 3.1-3.8 (Mixed Strategies through Finding Best Responses).

**Quiz**:    Quiz 1, in class.

## Week 3: Mixed Strategies and Rationalizability

### `r advdate(tue, 3)` (Lecture 5)

**Topic**:    Mixed strategies and the interpretation problem (Aumann; Harsanyi purification).    
**Reading**:    Notes, 3.9-3.12 (Finding Mixed Strategy Equilibria through What Is a Mixed Strategy?).

### `r advdate(thu, 3)` (Lecture 6)

**Topic**:    Rationalizability; common knowledge of rationality.    
**Reading**:    Notes, 4.1 (Rationalizability).

**Quiz**:    Quiz 2, in class.

## Week 4: Dynamic Games

### `r advdate(tue, 4)` (Lecture 7)

**Topic**:    Extensive form games; subgame perfection.    
**Reading**:    Notes, Ch 5 (Games and Time), 5.1-5.4 (Normal Form and Extensive Form through Subgame Perfect Equilibrium).

### `r advdate(thu, 4)` (Lecture 8)

**Topic**:    Backward induction; the centipede paradox.    
**Reading**:    Notes, 5.5 (Problems with Backwards Induction).

**Quiz**:    Quiz 3, in class.

## Week 5: Forward Induction and Bayesian Games

### `r advdate(tue, 5)` (Lecture 9)

**Topic**:    Forward induction. Van Damme's burning-money game; reasoning about what kind of player would have made an off-path move.    
**Reading**:    Notes, 5.6 (Money Burning Game).

### `r advdate(thu, 5)` (Lecture 10)

**Topic**:    Bayesian games and Bayesian Nash equilibrium. Types, incomplete information, and Nash equilibrium applied to type-conditional strategies.    
**Reading**:    Notes, Ch 6 (Bayesian Games), 6.1-6.5 (Two Kinds of Ignorance through Purification).

## Week 6: Signaling Games

### `r advdate(tue, 6)` (Lecture 11)

**Topic**:    Signaling games; perfect Bayesian equilibrium.    
**Reading**:    Notes, 6.6-6.8 (Perfect Bayesian Equilibrium through Signaling without Cooperation).

### `r advdate(thu, 6)` (Lecture 12)

**Topic**:    The beer-quiche game; the intuitive criterion.    
**Reading**:    Notes, 6.9 (The Intuitive Criterion). In-Koo Cho and David M. Kreps, "Signaling Games and Stable Equilibria", *Quarterly Journal of Economics* 102 (1987): 179-221. <https://www.jstor.org/stable/1885060>

**Quiz**:    Quiz 4, in class.

## Week 7: Asymmetric Information

### `r advdate(tue, 7)` (Lecture 13)

**Topic**:    Spence on job-market signaling.    
**Reading**:    Michael Spence, "Job Market Signaling", *Quarterly Journal of Economics* 87 (1973): 355-374. <https://www.jstor.org/stable/1882010>

### `r advdate(thu, 7)` (Lecture 14)

**Topic**:    Akerlof on lemons (rigorous treatment).    
**Reading**:    George A. Akerlof, "The Market for 'Lemons': Quality Uncertainty and the Market Mechanism", *Quarterly Journal of Economics* 84 (1970): 488-500. <https://www.jstor.org/stable/1879431>

## Week 8: Transition

### `r advdate(tue, 8)`

No class -- Fall Study Break (October 19-20).

### `r advdate(thu, 8)` (Lecture 15)

**Topic**:    Transition from strategic rationality to collective choice. Why aggregation is a different problem from strategy. Condorcet's paradox, now stated properly.    
**Reading**:    CCSW: Ch 1 (Introduction); Ch A1 (Enlightenment and Impossibility).    
**Recommended**:    Notes, Ch 7 (Group Decisions), which covers this ground quickly and informally before we slow down and do it properly.    
**Recommended**:    CCSW: New Introduction (2017). This is worth reading over the study break; it is the best short account of what the whole book is doing.

## Week 9: Foundations

### `r advdate(tue, 9)` (Lecture 16)

**Topic**:    The apparatus. Collective choice rules, the Pareto relation, and how little unanimity delivers. Preference relations, orderings and quasi-orderings, and the difference between a maximal set and a choice set, will be introduced at the board.    
**Reading**:    CCSW: Ch 2 (Unanimity); Ch 2* (Collective Choice Rules and Pareto Comparisons), section 2*1 only.    
**Recommended**:    CCSW: Ch 1* (Preference Relations). Keep this one open for the rest of the semester; it is the reference for notation we will use throughout. Section 2*2, on the Kaldor and Scitovsky compensation tests, is not needed for anything later in the course.

### `r advdate(thu, 9)` (Lecture 17)

**Topic**:    Arrow's conditions, and what each of them claims. May's theorem as the contrast case: with two options, anonymity, neutrality and positive responsiveness characterise majority rule. Then what goes wrong at three.    
**Reading**:    CCSW: Ch 3 (Collective Rationality).    
**Recommended**:    CCSW: Ch 3* (Social Welfare Functions); Ch 5 (Values and Choice); Ch 5* (Anonymity, Neutrality and Responsiveness).

**Essay**:    First essay due `r advdate(fri, 9)` at 5pm.

## Week 10: Arrow's Theorem, and the First Escape

### `r advdate(tue, 10)` (Lecture 18)

**Topic**:    Arrow's theorem: the proof, in Sen's decisive-set form. The spread of decisiveness, the contraction of decisive sets, and what the result does and does not establish. If there is time, what happens to the argument when the electorate is infinite.    
**Reading**:    CCSW: Ch A1* (Social Preference), to p. 288.    
**Recommended**:    CCSW: Ch 3* (Social Welfare Functions), closely.

### `r advdate(thu, 10)` (Lecture 19)

**Topic**:    First escape: restrict the domain. Single-peakedness and value restriction. Acyclicity will be introduced at the board. This closes the loop on the Condorcet case from Lecture 1.    
**Reading**:    CCSW: Ch 10 (Majority Choice and Related Systems).    
**Recommended**:    CCSW: Ch 10* (Restricted Preferences and Rational Choice).

**Quiz**:    Quiz 5, in class.

## Week 11: Two More Escapes

### `r advdate(tue, 11)` (Lecture 20)

**Topic**:    Second escape: weaken collective rationality. Quasi-transitivity, choice functions, social decision functions, the Pareto-extension rule, and what it costs (Gibbard's oligarchy result). The hinge: the spread of decisiveness needs only quasi-transitivity, while the contraction of decisive sets needs full transitivity, so we can see exactly which step of last week's proof fails.    
**Reading**:    CCSW: Ch A1* (Social Preference), pp. 289-293; Ch 4 (Choice Versus Orderings); Ch A2 (Rationality and Consistency).    
**Recommended**:    CCSW: Ch 4* (Social Decision Functions); Ch A2* (Problems of Social Choice). The rest of Ch A1*, on semi-orders and the topology of decisive sets, is past where we need to go, but it is where this line of results leads.

### `r advdate(thu, 11)` (Lecture 21)

**Topic**:    Third escape: enrich the information. Arrow's impossibility as a consequence of an ordinal, non-comparable informational base. What interpersonal comparison would have to mean, and whether it is available.    
**Reading**:    CCSW: Ch 7 (Interpersonal Aggregation and Comparability).    
**Recommended**:    CCSW: Ch 7* (Aggregation Quasi-Orderings); Ch 8 (Cardinality With or Without Comparability).

**Quiz**:    Quiz 6, in class.

## Week 12: Strategy and Liberty

### `r advdate(tue, 12)` (Lecture 22)

**Topic**:    Strategic manipulation. Voting rules as game forms; the Gibbard-Satterthwaite theorem; the Muller-Satterthwaite equivalence showing that it is Arrow's theorem in different clothes. The point in the course where the two halves meet.    
**Reading**:    Amartya Sen, "The Possibility of Social Choice", *American Economic Review* 89 (1999): 349-378. <https://www.aeaweb.org/articles?id=10.1257/aer.89.3.349>    
**Recommended**:    Allan Gibbard, "Manipulation of Voting Schemes: A General Result", *Econometrica* 41 (1973): 587-601. <https://www.jstor.org/stable/1914083>

### `r advdate(thu, 12)` (Lecture 23)

**Topic**:    A second impossibility: the Paretian liberal. Conflicts between the Pareto principle and minimal liberty.    
**Reading**:    CCSW: Ch 6 (Conflicts and Dilemmas); Ch 6* (The Liberal Paradox).    
**Recommended**:    Amartya Sen, "The Impossibility of a Paretian Liberal", *Journal of Political Economy* 78 (1970): 152-157. <https://doi.org/10.1086/259614>

**Quiz**:    Quiz 7, in class.

## Week 13: Rights

### `r advdate(tue, 13)` (Lecture 24)

**Topic**:    Rights and social choice. Sen's later response to his own paradox: goal rights systems, and the consequential evaluation of rights.    
**Reading**:    CCSW: Ch A5 (The Idea of Rights).    
**Recommended**:    CCSW: Ch A5* (Rights and Social Choice).

**Quiz**:    Quiz 8, in class. (Thursday is Thanksgiving recess, so this week's quiz is on Tuesday.)

### `r advdate(thu, 13)`

No class -- Thanksgiving recess (November 25-27).

## Week 14: Equity, Justice, Capability

### `r advdate(tue, 14)` (Lecture 25)

**Topic**:    Equity and justice. The weak equity axiom, and the limits of welfarism.    
**Reading**:    CCSW: Ch 9 (Equity and Justice).    
**Recommended**:    CCSW: Ch 9* (Impersonality and Collective Quasi-Orderings); Ch A3 (Justice and Equity).

### `r advdate(thu, 14)` (Lecture 26)

**Topic**:    The capability approach. What equality is equality of. Why Sen declines to specify a list of capabilities, and why that refusal is itself a position in social choice theory.    
**Reading**:    Amartya Sen, "Equality of What?", Tanner Lecture on Human Values, Stanford University, 22 May 1979; reprinted in *Choice, Welfare and Measurement*. <https://tannerlectures.org/lectures/equality-of-what/>    
**Recommended**:    CCSW: Ch A3* (Social Welfare Evaluation). Martha Nussbaum, "Capabilities as Fundamental Entitlements: Sen and Social Justice", *Feminist Economics* 9 (2003): 33-59. <https://doi.org/10.1080/1354570022000077926>

## Week 15: Democracy and Public Reasoning

### `r advdate(tue, 15)` (Lecture 27)

**Topic**:    Democracy and public engagement. Voting as one input among several, and the case for public reasoning as the mechanism that fixes the weights.    
**Reading**:    CCSW: Ch A4 (Democracy and Public Engagement).    
**Recommended**:    CCSW: Ch A4* (Votes and Majorities).

### `r advdate(thu, 15)` (Lecture 28)

**Topic**:    Reasoning and social decisions. Course wrap: how the two halves of the course fit together.    
**Reading**:    CCSW: Ch A6 (Reasoning and Social Decisions).    
**Recommended**:    CCSW: Ch 11 (Theory and Practice).

**FINAL ESSAY**: Due `r advdate(fri, 16)` at 5pm.

\newpage

# Plagiarism

Although team-work, and even co-authorship, is encouraged, plagiarism is strictly prohibited. You are responsible for making sure that none of your work is plagiarized. Be sure to cite work that you use, both direct quotations and paraphrased ideas. Any citation method that is tolerably clear is permitted, but if you'd like a good description of a citation scheme that works well in philosophy, look at <http://bit.ly/VDhRJ4>.

You are encouraged to discuss the course material, including assignments, with your classmates, but all written work that you hand in under your own name must be your own. If work is handed is as the work of two people, you are affirming that each person did a fair share of the work. (Note that when you're submitting work on Canvas, you have to each submit the paper, even if it is co-authored. That way Canvas knows that everyone has turned in work.)

You should also be familiar with the academic integrity policies of the College of Literature, Science & the Arts at the University of Michigan, which are available here: <http://www.lsa.umich.edu/academicintegrity/>. Violations of these policies will be reported to the Office of the Assistant Dean for Student Academic Affairs, and sanctioned with a course grade of F.

# Disability

The University of Michigan abides by the Americans with Disabilities Act of 1990, Section 504 of the Rehabilitation Act of 1973, and other applicable federal and state laws that prohibit discrimination on the basis of disability, which mandate that reasonable accommodations be provided for qualified students with disabilities.

If you have a disability, and may require some type of instructional and/or examination accommodation, please contact me early in the semester. If you have not already done so, you will also need to register with the Office of Services for Students with Disabilities. The office is located at G664 Haven Hall.

For more information on disability services at the University of Michigan, go to <http://ssd.umich.edu>.
