---
title: 'PHIL 305: Introduction to Formal Methods'
author: "Brian Weatherson"
date: "Winter 2026"
format:
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    fig-cap-location: bottom
    geometry: margin=1in
    mainfont: Atkinson Hyperlegible
    mathfont: STIX Two Math
    date-format: "[Winter] YYYY"
    include-in-header:
      text: |
        \linespread{1.05}
        \RedeclareSectionCommand[
          afterskip=0.3ex,
          beforeskip=0.8ex
          ]{subsection}
        \RedeclareSectionCommand[
          afterskip=0.1ex,
          beforeskip=0.5ex
          ]{subsubsection}
  html:
    theme: cosmo
    code-tools: false
    axe: true
    toc: true
    shift-heading-level-by: 1
    date-format: "[Winter] YYYY"
  docx: default
---

```{r}
#| label: setup
#| include: false
#| cache: false
#| message: false
#| warning: false

knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(fig.path = 'figs/')
knitr::opts_chunk$set(cache.path = '_cache/')
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
```

```{r}
#| label: dates_bib
#| include: false
#| cache: false

mon <- as.Date("2026-01-05") # Monday of the First Week
tue <- mon + 1
wed <- tue + 1
thu <- wed + 1
fri <- thu + 1

advdate <- function(obj, adv) {
 theday <- obj + 7*(adv-1)
 ttheday <- format(theday, format="%A, %B %d")
 return(ttheday)
}
```

| **Lead Instructor**: Brian Weatherson
| **Email**: weath@umich.edu
| **Web**: canvas.umich.edu
| **Office Hours**: Thursday 9.30-11.30 (Angell Hall, Rm 2207)
| **Discussion Section Leader**: Brett Thompson
| **Email**: brettdt@umich.edu
| **Office Hours**: Monday 4-6

&nbsp;

# Key Points

- The course has four parts:
  1. Propositional Logic;
  2. Modal Logic;
  3. Probability and Decision; and,
  4. Game Theory.
- Each part of the course has a different textbook.
- All the books are open access; they are freely available and linked through Canvas.
- There are 12 weekly assignments, due each week on Thursday at 5pm. The best 10 count for your final grade.
- There is a midterm test, in the last lecture before the break, and which counts for 10%.
- There is a final exam, which counts for 20%, and which covers all the material in the course.

# Course Description

This course will introduce some important formal tools that are used elsewhere in philosophy. We will look at propositional logic, the logic of modals and conditionals, probability theory and game theory. Obviously that's a lot to cover in a short time. The aim here is to make sure you understand the basics, and the symbolism, so you can follow simple applications of these tools, and you have the foundations to understand more complicated applications.

# Required Materials

There are four textbooks for the course. All of them are open access, and hence free. The second is on Canvas, the other three have to be downloaded from elsewhere.

- _forall x: Calgary_ by P. D. Magnus, Tim Button, J. Robert Loftis, Robert Trueman, Aaron Thomas-Bolduc and Richard Zach. Available at <http://forallx.openlogicproject.org>.
- _Boxes and Diamonds: Ann Arbor remix_ by Richard Zach and edited by Brian Weatherson. Available on Canvas.
- _Odds and Ends_ by Jonathan Weisberg. Available at <https://jonathanweisberg.org/vip/>.
- _Game Theory_ by Giacomo Bonnano. Available at <https://faculty.econ.ucdavis.edu/faculty/bonanno/GT_Book.html>.

# Course Requirements

- There will be 12 weekly assignments. Of these your best 10 will count towards your grade, with each counting for about 7%.
- There will be a midterm exam which counts for 10% of the grade.
- There will also be a final exam that counts for 20% of the grade.

# Summary of Grading System

1. Weekly assignments - 7% each, 10 assignments count, 70% total.
2. Midterm exam - 10%.
2. Final exam - 20%.

# Canvas

There is a Canvas site for this course, which can be accessed from <https://canvas.umich.edu>. Course documents (syllabus, lecture notes, assignments) will be available from this site. Please make sure that you can access this site. Consult the site regularly for announcements, including changes to the course schedule. And there are many tools on the site to communicate with each other, and with me.

\pagebreak

# Class Schedule

The readings will all be from the four textbooks. You should do the reading before the scheduled class.

## Week 1: Introduction

### `r advdate(wed, 1)`

**Topic**:    Introduction  
**Reading**:    _forall x_, Chapters 1-2.

## Week 2: Symbols and Truth Tables

### `r advdate(mon, 2)`

**Topic**:    Symbolization  
**Reading**:    _forall x_, Chapters 4-6.

### `r advdate(wed, 2)`

**Topic**:    Truth Tables  
**Reading**:    _forall x_, Chapters 9-11.  
**Assignment**:    Assignment 1 due `r advdate(thu, 2)` at 5pm.

## Week 3: Validity

### `r advdate(mon, 3)`

No class for MLK Day

### `r advdate(wed, 3)`

**Topic**:    Truth Tables and Validity  
**Reading**:    _forall x_, Chapter 12.  
**Assignment**:  Assignment 2 due `r advdate(thu, 3)` at 5pm.

## Week 4: Trees

### `r advdate(mon, 4)`

**Topic**:    Truth Trees  
**Reading**:    _Boxes and Diamonds_, Sections 2.1-2.3  

### `r advdate(wed, 4)`

No lecture, because I'm away at a conference. There will be a discussion section.  
**Assignment**:  Assignment 3 due `r advdate(thu, 4)` at 5pm.

## Week 5: Trees and Modality

### `r advdate(mon, 5)`

**Topic**:    Using Truth Trees  
**Reading**:    _Boxes and Diamonds_, Sections 2.4-2.5  

### `r advdate(wed, 5)`

**Topic**:    Varieties of Modality  
**Reading**:    _Boxes and Diamonds_, sections 3.1-3.3.  
**Assignment**:    Assignment 4 (on trees) due `r advdate(thu, 4)` at 5pm.

## Week 6: Validity in Modal Logic

### `r advdate(mon, 6)`

**Topic**:    Models and Frames  
**Reading**:    _Boxes and Diamonds_, sections 3.4-4.5.  

### `r advdate(wed, 6)`

**Topic**:    Modal Tableau  
**Reading**:    _Boxes and Diamonds_, chapter 5.  
**Assignment**:    Assignment 5 due `r advdate(thu, 6)` at 5pm.

## Week 7: Conditionals

### `r advdate(mon, 7)`

**Topic**:    Examples of Modal Tableau  
**Reading**:    No new reading  

### `r advdate(wed, 7)`

**Topic**:    Varieties of Conditionals   
**Reading**:    _Boxes and Diamonds_, chapter 6.  
**Assignment**:    Assignment 6 (on modal tableau) due `r advdate(thu, 7)` at 5pm.

## Week 8: Counterfactuals

### `r advdate(mon, 7)`

**Topic**:    Counterfactual Conditionals  
**Reading**:    _Boxes and Diamonds_, chapter 7.

### `r advdate(wed, 8)`

**MIDTERM EXAM**  

**Assignment**:    Assignment 7 due `r advdate(thu, 8)` at 5pm.

## Week 9: Starting on Probability

### `r advdate(mon, 10)`

**Topic**:    Probability Basics  
**Reading**:    _Odds and Ends_, chapters 1 and 5

### `r advdate(wed, 10)`

**Topic**:    Probability Trees  
**Reading**:    _Odds and Ends_, chapter 1  
**Assignment**:    No assignment this week.

## Week 10: Conditional Probability

### `r advdate(mon, 11)`

**Topic**:    Conditional Probability  
**Reading**:    _Odds and Ends_, chapter 6

### `r advdate(wed, 11)`

**Topic**:    Inverting Conditional Probability  
**Reading**:    _Odds and Ends_, chapter 8  
**Assignment**:    Assignment 8 due `r advdate(thu, 11)` at 5pm.

## Week 11: Probability and Decision

### `r advdate(mon, 12)`

**Topic**:    Expected Utility  
**Reading**:    _Odds and Ends_, chapters 11 and 12

### `r advdate(wed, 12)`

**Topic**:    Utility and Money  
**Reading**:    _Odds and Ends_, sections 12.5 and 13.1.  
**Assignment**:    Assignment 9 due `r advdate(thu, 12)` at 5pm.

\pagebreak

## Week 12: What are Games?

### `r advdate(mon, 13)`

**Topic**:    Games, Payouts and Utilities  
**Reading**:    _Game Theory_, sections 2.1 and 2.2.

### `r advdate(wed, 13)`

**Topic**:    Dominant Strategies and Equilibria  
**Reading**:    _Game Theory_, sections 2.5 and 2.6.  
**Assignment**:    Assignment 10 due `r advdate(thu, 13)` at 5pm.

## Week 13: Games and Time

### `r advdate(mon, 14)`

**Topic**:    Dynamic Games  
**Reading**:    _Game Theory_, sections 3.1-3.3.

### `r advdate(wed, 14)`

**Topic**:    Backward Induction  
**Reading**:    _Game Theory_, sections 3.4 and 3.5.  
**Assignment**:    Assignment 11 due `r advdate(thu, 14)` at 5pm.

## Week 14: Games with Cardinal Payoffs

### `r advdate(mon, 15)`

**Topic**:    Nash Equilibria  
**Reading**:    _Game Theory_, sections 6.1‑6.3.

### `r advdate(wed, 15)`

**Topic**:    Signaling Games  
**Reading**:    No new reading  
**Assignment**:    Assignment 12 due `r advdate(thu, 15)` at 5pm.

There will be a revision class on `r advdate(mon, 16)`, with no new material, to go over the sample exam.

# Extensions

Because we will release the answers to the quizzes at the due date, we won't be allowing extensions on them.

That's why we're only counting the best 10 out of 12 assignments. People who are sick, or have other obligations, can skip up to two assignments without penalty.

# Plagiarism

Although team-work, and even co-authorship, is encouraged, plagiarism is strictly prohibited. You are responsible for making sure that none of your work is plagiarized. Be sure to cite work that you use, both direct quotations and paraphrased ideas. Any citation method that is tolerably clear is permitted, but if you'd like a good description of a citation scheme that works well in philosophy, look at <http://bit.ly/VDhRJ4>.

You are encouraged to discuss the course material, including assignments, with your classmates, but all written work that you hand in under your own name must be your own. If work is handed is as the work of two people, you are affirming that each person did a fair share of the work. (Note that when you're submitting work on Canvas, you have to each submit the paper, even if it is co-authored. That way Canvas knows that everyone has turned in work.)

You should also be familiar with the academic integrity policies of the College of Literature, Science & the Arts at the University of Michigan, which are available here: <http://www.lsa.umich.edu/academicintegrity/>. Violations of these policies will be reported to the Office of the Assistant Dean for Student Academic Affairs, and sanctioned with a course grade of F.

# Disability

The University of Michigan abides by the Americans with Disabilities Act of 1990, Section 504 of the Rehabilitation Act of 1973, and other applicable federal and state laws that prohibit discrimination on the basis of disability, which mandate that reasonable accommodations be provided for qualified students with disabilities.

If you have a disability, and may require some type of instructional and/or examination accommodation, please contact me early in the semester. If you have not already done so, you will also need to register with the Office of Services for Students with Disabilities. The office is located at G664 Haven Hall.

For more information on disability services at the University of Michigan, go to <http://ssd.umich.edu>.


