---
title: 'PHIL 342: Introduction to Groups and Choices'
author: "Brian Weatherson"
date: "Winter 2026"
format:
  pdf:
    keep-tex: true
    pdf-engine: xelatex
    fig-cap-location: bottom
    geometry: margin=1in
    mainfont: Atkinson Hyperlegible
    mathfont: STIX Two Math
    date-format: "[Winter] YYYY"
    include-in-header:
      text: |
        \linespread{1.05}
        \RedeclareSectionCommand[
          afterskip=0.3ex,
          beforeskip=0.8ex
          ]{subsection}
        \RedeclareSectionCommand[
          afterskip=0.1ex,
          beforeskip=0.5ex
          ]{subsubsection}
  html:
    theme: cosmo
    code-tools: false
    axe: true
    toc: true
    shift-heading-level-by: 1
    date-format: "[Winter] YYYY"
  docx: default
---

```{r}
#| label: setup
#| include: false
#| cache: false
#| message: false
#| warning: false

knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(fig.path = 'figs/')
knitr::opts_chunk$set(cache.path = '_cache/')
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
```

```{r}
#| label: dates_bib
#| include: false
#| cache: false

mon <- as.Date("2026-01-05") # Monday of the First Week
tue <- mon + 1
wed <- tue + 1
thu <- wed + 1
fri <- thu + 1

advdate <- function(obj, adv) {
 theday <- obj + 7*(adv-1)
 ttheday <- format(theday, format="%A, %B %d")
 return(ttheday)
}
```

| **Lead Instructor**: Brian Weatherson
| **Email**: weath@umich.edu
| **Web**: canvas.umich.edu
| **Office Hours**: Thursday 9.30-11.30 (Angell Hall, Rm 2207)
| **Discussion Section Leader**: Mitch Barrington
| **Email**: umitch@umich.edu
| **Office Hours**: Monday 1-3 (Angell Hall, Rm 2253)

&nbsp;

# Key Points

- This course examines philosophical questions that arise when individual choices interact.
- The course has two main parts: pooling (how we form collective attitudes from individual attitudes) and modeling (what simple models reveal about social structures).
- We'll study voting methods, preference aggregation, opinion pooling, game theory, signaling, and models of scientific communities.
- Three textbooks are available electronically through the library: *Opinion Pooling* by Lee Elkin and Richard Pettigrew, *Modeling Scientific Communities* by Cailin O'Connor, and *Essentials of Game Theory* by Kevin Leyton-Brown and Yoav Shoham.
- Assessment consists of four short assignments (10% each), two essays (25% each), and discussion section participation (10%)

# Course Description

This course introduces topics at the intersection of philosophy, game theory, and social choice theory. We examine how individual choices and judgments combine to form group decisions, and what mathematical models can teach us about social phenomena. 

The first half focuses on pooling problems: when we try to aggregate individual beliefs or preferences into collective judgments, we encounter surprising paradoxes and impossibility results. We'll explore voting methods, Arrow's theorem, judgment aggregation, and methods for pooling probabilistic opinions. 

The second half looks at the use mathematical and computational models to understand social dynamics. The topics we'll cover include focal points, signaling in markets, iterated games, and agent-based models of scientific communities and knowledge production.

# Required Materials

There are three textbooks for the course. All of them are available electronically through the library (and the first is open access, so doesn't even need a library login).

- Lee Elkin and Richard Pettigrew, [Opinion Pooling](https://www.cambridge.org/core/elements/opinion-pooling/40647AA21F276CF80F67398058682294)
- Cailin O'Connor, [Modeling Scientific Communities](https://www.cambridge.org/core/elements/abs/modelling-scientific-communities/1ED3515216067E40A37A72094EE3CB15) (if you are off-campus you can access this via [the library](https://search.lib.umich.edu/catalog/record/99187887203306381))
- Kevin Leyton-Brown and Yoav Shoham, [Essentials of Game Theory](https://link-springer-com.proxy.lib.umich.edu/book/10.1007/978-3-031-01545-8)

# Course Requirements

- There will be 4 short assignments, each worth 10% of the grade
- There will be 2 essays, each worth 25% of the grade.
- Lecture participation will not be graded, and we won't take attendance here. But 10% of the grade is for discussion section participation (and any other assignments the discussion section leader chooses to include).

# Summary of Grading System

1. Short assignments - 4 assignments, 10% each: 40%
2. Essays - 2 essays, 25% each: 50%
3. Discussion section participation. (10%)

# Canvas

There is a Canvas site for this course, which can be accessed from <https://canvas.umich.edu>. Course documents will be available from this site. Please make sure that you can access this site. Consult the site regularly for announcements, including changes to the course schedule. And there are many tools on the site to communicate with each other, and with me.

There is also a github site for the course which contains more documents, and it is at <https://bweatherson.github.io/W26-Phil342-site/>.

\pagebreak

# Class Schedule

Readings should be done before the scheduled class.

## Week 1: Introduction

### `r advdate(wed, 1)`

**Topic**:    Introduction  
**Reading**:    None.

**NOTE**: No discussion section this week.

## Week 2: Voting

### `r advdate(mon, 2)`

**Topic**:    Voting Methods    
**Reading**:    Eric Pacuit, [Voting Methods](https://plato.stanford.edu/entries/voting-methods/)

### `r advdate(wed, 2)`

**Topic**:    Arrow's Theorem    
**Reading**:    Michael Morreau, [Arrow's Theorem](https://plato.stanford.edu/archives/win2022/entries/arrows-theorem/) 

## Week 3: Arrow's Theorem

### `r advdate(mon, 3)`

No class for MLK Day

### `r advdate(wed, 3)`

**Topic**:    Responses to Arrow  
**Reading**:    No new reading

## Week 4: Sen's Paradox

### `r advdate(mon, 4)`

**Topic**:    Sen's Paradox  
**Reading**:    Amartya Sen, [The Impossibility of a Paretian Liberal](https://www.jstor.org/stable/1829633?seq=1)  
**Suggested Reading**:   Amartya Sen, [Liberty and Social Choice](https://www.jstor.org/stable/2026284?seq=1)

### `r advdate(wed, 4)`

No lecture, because I'm away at a conference. There will be a discussion section.  

**Assignment**:    First short assignment due `r advdate(thu, 4)` at 5pm.

## Week 5: Judgment Aggregation

### `r advdate(mon, 5)`

**Topic**:    From Belief to Probability   
**Reading**:    _Opinion Pooling_, Ch. 1 and section 2.1.   
**Suggested Reading**:

- David Christensen, [Epistemology of Disagreement: The Good News](https://philpapers.org/go.pl?id=CHREOD&proxyId=&u=https%3A%2F%2Fphilpapers.org%2Farchive%2FCHREOD.pdf)
- Adam Elga, [Reflection and Disagreement](https://philsci-archive.pitt.edu/2940/1/refdis.pdf)
- Thomas Kelly, [The Epistemic Significance of Disagreement](https://philpapers.org/go.pl?id=KELTES-3&proxyId=&u=https%3A%2F%2Fwww.princeton.edu%2F~tkelly%2Fesod.pdf)

### `r advdate(wed, 5)`

**Topic**:    Reasons for Aggregation   
**Reading**:    _Opinion Pooling_, Rest of chapter 2.  
**Suggested Reading**:

- Roger White, [Epistemic Permissiveness](https://philpapers.org/go.pl?id=WHIEP&proxyId=&u=http%3A%2F%2Fweb.mit.edu%2Frog%2Fwww%2Fpapers%2FEP.pdf)
- Miriam Schoenfield, [Permission to Believe: Why Permissivism Is True and What It Tells Us About Irrelevant Influences on Belief](https://philpapers.org/go.pl?id=SCHPTB&proxyId=&u=https%3A%2F%2Fphilpapers.org%2Farchive%2FSCHPTB.pdf)

## Week 6: Constraints on Aggregation

### `r advdate(mon, 6)`

**Topic**:    Motivating the Constraints  
**Reading**:    _Opinion Pooling_, Chapter 3.

### `r advdate(wed, 6)`

**Topic**:    Linear and Geometric Pooling    
**Reading**:    _Opinion Pooling_, section 4.1.  
**Suggested Reading**:

- John Hawthorne, Lara Buchak, and Jeffrey Sanford Russell, [Groupthink](https://philpapers.org/go.pl?id=RUSG&proxyId=&u=https%3A%2F%2Fphilpapers.org%2Farchive%2FRUSG.pdf)
- Jan-Willem Romeijn, [An Interpretation of Weights in Linear Opinion Pooling](https://philpapers.org/go.pl?id=ROMAIO&proxyId=&u=https%3A%2F%2Fdx.doi.org%2F10.1017%2Fepi.2020.45)
- Jean Baccelli & Rush T. Stewart, [Support for Geometric Pooling](https://philpapers.org/go.pl?id=BACSFG&proxyId=&u=https%3A%2F%2Fphilpapers.org%2Farchive%2FBACSFG.pdf)

## Week 7: Evaluating the Strategies

### `r advdate(mon, 7)`

**Topic**:    Multiplicative Pooling  
**Reading**:    _Opinion Pooling_, sections 4.2-4.3

### `r advdate(wed, 7)`

**Topic**:    Group Rationality   
**Reading**:    _Opinion Pooling_, chapter 5.  
**Suggested Reading**:

- James Joyce, [A Nonpragmatic Vindication of Probabilism](https://philpapers.org/go.pl?id=JOYANV-2&proxyId=&u=http%3A%2F%2Ffitelson.org%2Ftopics%2Fjoyce_1998.pdf)
- Rush T. Stewart & Ignacio Ojea Quintana, [Learning and Pooling, Pooling and Learning](https://philpapers.org/go.pl?id=STELAP-15&proxyId=&u=https%3A%2F%2Fdx.doi.org%2F10.1007%2Fs10670-017-9894-2)

**Assignment**:    Second short assignment due `r advdate(thu, 7)` at 5pm.

## Week 8: Schelling

### `r advdate(mon, 8)`

**Topic**:    Focal Points  
**Reading**:    Judith Mehta, Chris Starmer, and Robert Sugden, [The Nature of Salience: An Experimental Investigation of Pure Coordination Games](https://www.jstor.org/stable/2118074)  
**Suggested Reading**:    Robert Sugden, [A Theory of Focal Points](https://www.jstor.org/stable/2235016)

### `r advdate(wed, 8)`

**Topic**:    Segregation  
**Reading**:    Vi Hart and Nicky Case, [Parable of the Polygons](https://ncase.me/polygons/)  
**Suggested Reading**:    Thomas Schelling, [Dynamic Models of Segregation](https://www.stat.berkeley.edu/~aldous/157/Papers/Schelling_Seg_Models.pdf)

**Essay**:    First essay due `r advdate(thu, 8)` at 5pm.

## Week 9: Game Theory

### `r advdate(mon, 10)`

**Topic**:    One-shot games  
**Reading**:    _Essentials of Game Theory_, Chapters 1 and 2

### `r advdate(wed, 10)`

**Topic**:    Dynamic games  
**Reading**:    _Essentials of Game Theory_, Chapters 4 and 5  

## Week 10: Axelrod

### `r advdate(mon, 11)`

**Topic**:    Iterated Prisoners' Dilemma    
**Reading**:    Robert Axelrod and William D. Hamilton, [The Evolution of Cooperation](https://websites.umich.edu/~axe/research/Axelrod%20and%20Hamilton%20EC%201981.pdf)   
**Suggested reading**:     Robert Axelrod, [Effective Choice in the Prisoner's Dilemma](https://esp.mit.edu/download/2dd80247-be60-4e73-b49d-91b5b6988ff0/S12850_axelrod_1980.pdf)

### `r advdate(wed, 11)`

**Topic**:    Modern Approaches to IPD   
**Reading**:    Alexander J. Stewart and Joshua B. Plotkin, [Extortion and cooperation in the Prisoner’s Dilemma](https://www.pnas.org/doi/10.1073/pnas.1208087109)    
**Suggested Reading**:    William H. Press and Freeman J. Dyson, [Iterated Prisoner’s Dilemma contains strategies that dominate any evolutionary opponent](https://www.pnas.org/doi/abs/10.1073/pnas.1206569109)    
**Assignment**:    Third short assignment due `r advdate(thu, 11)` at 5pm.

## Week 11: Signals

### `r advdate(mon, 12)`

**Topic**:    Signals in car markets   
**Reading**:    George Akerlof, [The Market for Lemons](https://www.jstor.org/stable/1879431)

### `r advdate(wed, 12)`

**Topic**:    Signals in employment markets   
**Reading**:    Michael Spence, [Signaling in Retrospect and the Informational Structure of Markets](https://www.nobelprize.org/uploads/2018/06/spence-lecture.pdf)   
**Suggested Reading**:    Michael Spence, [Job Market Signaling](https://www.jstor.org/stable/1882010)


## Week 12: Social Networks and Scientific Knowledge 

### `r advdate(mon, 13)`

**Topic**:    Models of evidence sharing   
**Reading**:    

- _Modeling Scientific Communities_, Chapter 4 through 4.2.1.
- Kevin J. S. Zollman, [The Communication Structure of Epistemic Communities](https://doi.org/10.1086/525605)

### `r advdate(wed, 13)`

**Topic**:    Evaluating models of evidence sharing    
**Reading**:    _Modeling Scientific Communities_, rest of Chapter 4.     
**Suggested Reading**:    Sarita Rosenstock, Justin Bruner, and Cailin O’Connor, [In Epistemic Networks, Is Less Really More?](https://cailinoconnor.com/wp-content/uploads/2015/03/In_Epistemic_Networks_Is_Less_Really_Mor-FINAL-VERSION.pdf).

## Week 13: The Credit Economy

### `r advdate(mon, 14)`

**Topic**:    The Matthew Effect     
**Reading**:    

- _Modeling Scientific Communities_, Chapter 2 through 2.6. (Section 2.6 is the most important part.)
- Robert K. Merton, [The Matthew Effect in Science](https://www.science.org/doi/10.1126/science.159.3810.56)

### `r advdate(wed, 14)`

**Topic**:    Credit and Bias     
**Reading**:    

- _Modeling Scientific Communities_, rest of Chapter 2.
- Remco Heesen, [Academic Superstars: Competent or Lucky?](https://link.springer.com/article/10.1007/s11229-016-1146-5) 

**Suggested Reading**:    Hannah Rubin, [Structural causes of citation gaps](https://link.springer.com/article/10.1007/s11098-021-01765-3)    
**Assignment**:    Fourth short assignment due `r advdate(thu, 14)` at 5pm.

## Week 14: Epistemic Landscapes

### `r advdate(mon, 15)`

**Topic**:    Landscapes and exploration  
**Reading**:    

- _Modeling Scientific Communities_, Chapter 5 through 5.1.
- Michael Weisberg and Ryan Muldoon, [Epistemic Landscapes and the Division of Cognitive Labor](https://www.jstor.org/stable/10.1086/644786?seq=1)

### `r advdate(wed, 15)`

**Topic**:    Rugged epistemic landscapes  
**Reading**:    

- _Modeling Scientific Communities_, rest of chapter 5.
- Johanna Thoma, [The Epistemic Division of Labor Revisited](https://eprints.lse.ac.uk/88129/1/Thoma_Division%20of%20Labour_Accepted.pdf)

There will be a revision class on `r advdate(mon, 16)`, with no new material, to go over any questions people have about the final essay.

**FINAL ESSAY**: Due `r advdate(tue, 16)` at 5pm.

\newpage

# Plagiarism

Although team-work, and even co-authorship, is encouraged, plagiarism is strictly prohibited. You are responsible for making sure that none of your work is plagiarized. Be sure to cite work that you use, both direct quotations and paraphrased ideas. Any citation method that is tolerably clear is permitted, but if you'd like a good description of a citation scheme that works well in philosophy, look at <http://bit.ly/VDhRJ4>.

You are encouraged to discuss the course material, including assignments, with your classmates, but all written work that you hand in under your own name must be your own. If work is handed is as the work of two people, you are affirming that each person did a fair share of the work. (Note that when you're submitting work on Canvas, you have to each submit the paper, even if it is co-authored. That way Canvas knows that everyone has turned in work.)

You should also be familiar with the academic integrity policies of the College of Literature, Science & the Arts at the University of Michigan, which are available here: <http://www.lsa.umich.edu/academicintegrity/>. Violations of these policies will be reported to the Office of the Assistant Dean for Student Academic Affairs, and sanctioned with a course grade of F.

# Disability

The University of Michigan abides by the Americans with Disabilities Act of 1990, Section 504 of the Rehabilitation Act of 1973, and other applicable federal and state laws that prohibit discrimination on the basis of disability, which mandate that reasonable accommodations be provided for qualified students with disabilities.

If you have a disability, and may require some type of instructional and/or examination accommodation, please contact me early in the semester. If you have not already done so, you will also need to register with the Office of Services for Students with Disabilities. The office is located at G664 Haven Hall.

For more information on disability services at the University of Michigan, go to <http://ssd.umich.edu>.


